Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-15 write-ups/PLDI/paper.tex
Related work progress.

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multi-tasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement, due to the need for multi-tasking and energy throttling,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
retain the essence of the occupancy-bound model, which is
simple to implement efficiently, supporting preemption
only in key special cases.

@@

\item Our experiments with a range of applications using three GPUs show that the model can enable efficient multi-tasking of cooperative and non-cooperative
  tasks.

@@

The semantics presented in \mysec\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.
%
However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient} execution of algorithms that require fair scheduling, and
\emph{responsive} multi-tasking, so that the GPU can be shared between
cooperative kernels and other shorter tasks with soft real-time constraints.
%
To achieve this balance, an implementation of the cooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet the following non-functional requirements.

@@

\hline
\end{tabular}\\
{\footnotesize
$^{\ast}$color ecology, $^{\dagger}$bc 1k\_128k, $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
}
\caption{Execution time slowdown associated with cooperative scheduling mode without multi-tasking}
\label{tab:overhead}
\end{table}

@@

As we have demonstrated for a representative set of irregular
algorithms, our cooperative kernels model allows blocking algorithms
to be upgraded to run under a fair scheduler in a manner that, with
appropriate care, facilitates responsive multi-tasking.

@@

Our solution does not impose a performance penalty on non-cooperative
kernels, which run in an unmodified manner, and careful placement of
our new language primitives that exploit the structure of a
cooperative kernel may enable more efficient multi-tasking than
kernel-oblivious preemption.

\paragraph{GPU multitasking}
%
\cite{DBLP:journals/tog/SteinbergerKBKDS14} \ADComment{Shows how to
  keep a GPU busy by flooding it with work.  Would absolutely not
  stand up to preemption.}

\cite{DBLP:conf/usenix/KatoLRI11} \ADComment{Idea seems to be to
  accept that GPUs do not offer epreemption, but find a reasonable way
  to schedule tasks to maximise responsiveness or throughput.}

\paragraph{Preemption support for GPUs}
%
\cite{DBLP:conf/asplos/ParkPM15} \ADComment{Investigates how one might
  implement preemption, i.e.\ shows that it can be achieved.
  Simulator only preemption.  Not sure whether their approach would
  give fair scheduling.}
%
Proposals to implement preemption, but these require hardware support and may be overkill.  Advantage: fully automatic; disadvantage: doesn't capture domain knowledge.
%
\ADComment{Kepler preemption support?  Mention in related work?}

@@

\cite{DBLP:conf/ics/WuCLSV15} \ADComment{Basic idea: map tasks to
  SMs, not to workgroups, so that tasks that would benefit from being
  processed by the same SM can be scheduled to the same SM.  Has
  little to do with preemption, but more to do with irregular
  computation.}

>>>>

Despite its practical prevalence, none of the current GPU programming
models actually mandate occupancy-bound execution.  Further, there are
clear reasons why this model is undesirable.
% from the point-of-view of
%GPU vendors and end users.
First, the execution model does not enable
multitasking, because a workgroup effectively \emph{owns} a compute
unit until the workgroup has completed execution.  If a long-running
kernel occupies all compute units, the GPU cannot be used for other
tasks, such as graphics rendering, in the meantime.
%Precisely because
%of this problem, operating systems employ a GPU \emph{watchdog} to
%heuristically detect and kill long-running computations on the GPU
%being used for display rendering~\cite{DBLP:conf/iwocl/SorensenD16}.
Second, \emph{energy throttling} is
an important concern for battery-powered devices, such as smartphones
and tablets~\cite{DBLP:journals/comsur/Vallina-RodriguezC13}.  In the future, it will be desirable for a mobile GPU driver to be able to
power down one or more compute units if the battery level is low or if
the user puts the device into a low power state.  Implementing energy
throttling would require being able to suspend the execution of a
workgroup, power down the compute unit to which the workgroup was
bound, and schedule the workgroup for later execution on a different
compute unit.

Our assessment, informed by discussions with a number of industrial
practitioners who have been involved in the OpenCL and/or HSA
standardisation efforts
(including~\cite{PersonalCommunicationRichards,PersonalCommunicationHowes}
and various engineers from GPU vendors), is that (1) GPU vendors will
not commit to the occupancy-bound execution model that they currently
implement, due to the need for multitasking and energy throttling,
yet (2) GPU vendors will not guarantee fair scheduling, due to the
runtime overhead associated with supporting full preemption of
workgroups~\cite{ISCAPAPERSreeMentioned}.  Vendors instead wish to
retain the essence of the occupancy-bound model, which is
simple to implement efficiently, supporting preemption
only in key special cases.

@@

\item Our experiments with a range of applications using three GPUs show that the model can enable efficient multitasking of cooperative and non-cooperative
  tasks.

@@

The semantics presented in \mysec\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.
%
However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient} execution of algorithms that require fair scheduling, and
\emph{responsive} multitasking, so that the GPU can be shared between
cooperative kernels and other shorter tasks with soft real-time constraints.
%
To achieve this balance, an implementation of the cooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet the following non-functional requirements.

@@

\hline
\end{tabular}\\
{\footnotesize
$^{\ast}$color ecology, $^{\dagger}$bc 1k\_128k, $^{\ddagger}$sssp usa\_ny, $^{\diamond}$octree
}
\caption{Execution time slowdown associated with cooperative scheduling mode without multitasking}
\label{tab:overhead}
\end{table}

@@

As we have demonstrated for a representative set of irregular
algorithms, our cooperative kernels model allows blocking algorithms
to be upgraded to run under a fair scheduler in a manner that, with
appropriate care, facilitates responsive multitasking.

@@

Our solution does not impose a performance penalty on non-cooperative
kernels, which run in an unmodified manner, and careful placement of
our new language primitives that exploit the structure of a
cooperative kernel may enable more efficient multitasking than
kernel-oblivious preemption.

\paragraph{Optimizing GPU scheduling}
%
A number of works have considered how best to schedulel dynamic
workloads on GPUs.  The Whippletree approach employs a persistent
megakernel to schedule multiple, dynamic tasks in a manner that aims
to best utilise GPU
resources~\cite{DBLP:journals/tog/SteinbergerKBKDS14}.  The TimeGraph
approach similarly aims to optimise scheduling of competing workloads,
in the context of OpenGL, using GPU-CPU synchronisation so that the
CPU can monitor and control the work that is scheduled on the
GPU~\cite{DBLP:conf/usenix/KatoLRI11}.  To best utilise the compute
units of a GPU for a particular kernel, SM-centric program
transformations have been proposed, which aim to direct work to
specific compute units, rather than to specific workgroups, based on
the observation that in practice multiple workgroups may share the
same compute unit, so that said workgrops might be in a good position
to exploit memory caches associated with that compute
unit~\cite{DBLP:conf/ics/WuCLSV15}.  None of these approaches tackle
the problem of \emph{fair} scheduling on GPUs that our work addresses, thus they do not aid in safe deployment of blocking irregular algorithms.

%
\ADComment{Kepler preemption support?  Mention in related work?}

@@

\cite{} \ADComment{Basic idea: map tasks to
  SMs, not to workgroups, so that tasks that would benefit from being
  processed by the same SM can be scheduled to the same SM.  Has
  little to do with preemption, but more to do with irregular
  computation.}
